---
title: "Data 607 Assignment 3"
author: "Steven Gonzalez"
date: "9/22/2024"
output: html_document
---

## Overview / Introduction
In this assignment we explore the uses of normalization and character manipulation in data frames. We start off by providing at least three data frames in R that demonstrate normalization. We then use the 173 majors listed in fivethirtyeight.com’s College Majors [data set](https://raw.githubusercontent.com/fivethirtyeight/data/refs/heads/master/college-majors/majors-list.csv) to provide code that identifies the majors that contain either "DATA" or "STATISTICS." To finish off, we handle a few exercises revolving around regular expressions.

### Load Packages
As always, let's start off by loading the necessary packages.
```{r load, message = FALSE}
library(tidyverse)
library(dplyr)
```

### Load Data
For the data frames, I decided to recreate a data set used on the "[Decomplexify](https://www.youtube.com/@decomplexify)" YouTube channel in a video entitled "[Learn Database Normalization - 1NF, 2NF, 3NF, 4NF, 5NF](https://www.youtube.com/watch?v=GFQaEYEc8_8&t=1349s)." After recreating the data set and making it available on GitHub, we then have to load it into our environment.
```{r load video game data}
url <- url('https://raw.githubusercontent.com/Stevee-G/Data607/refs/heads/main/Video_Game_Data.csv')
video_game <- read.csv(url)
video_game
```

### Normalize Data
Now, we can go ahead and recreate the normalized data frames that were exemplified in the video. The `players` and `players_inventory` data frames are easily constructed from the original data set. However, the `layer_skill_levels` data frame must be built from scratch using the relationship between `Player_Skill_Level` and `Player_Rating` that was shown in the video. Once all is said and done, we can see that all three data frames have just one instance per row and each cell is in atomic level. They also abide by the rules surrounding non-key attributes as everyone one in a table key, the whole key, and nothing but the key.
```{r normalize video game data}
players <- subset(video_game,select = c(Player_ID,Player_Rating)) %>% 
  distinct()
players
player_inventory <- subset(video_game,select = c(Player_ID,Item_Type,Item_Quantity)) %>% 
  distinct()
player_inventory
player_skill_levels <- data.frame(Player_Skill_Level = 1:9) %>% 
  mutate(Player_Rating = case_when(
    Player_Skill_Level == 1:3 ~ "Beginner",
    Player_Skill_Level == 4:6 ~ "Intermediate",
    Player_Skill_Level == 7:9 ~ "Advanced"))
player_skill_levels
```

### Load FiveThirtyEight’s College Majors Data
Next, we will load the data used in fivethirtyeight.com's "[College Majors](https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/)" article.
```{r }
url2 <-
  url('https://raw.githubusercontent.com/fivethirtyeight/data/refs/heads/master/college-majors/majors-list.csv')
college_majors <- read.csv(url2)
head(college_majors)
```

### Identify Requested Majors
We can now search for the majors that contain either "DATA" or "STATISTICS" within the `college_majors` data set using the `filter()` and `grepl()` functions.
```{r tidy and transform}
college_majors %>% filter(grepl('DATA|STATISTICS', Major))
```

### Expressions Practice
Next we've been tasked with describing what the expressions below will match:

(.)\1\1 - This expression will not match anything since it is not contained within quotation marks and the back references are missing a backslash.

"(.)(.)\\2\\1" - This expression will match any string where a pair of characters is followed by the same pair in reverse order.

(..)\1 - This expression will not match anything since it is not contained within quotation marks and the back reference is missing a backslash.

"(.).\\1.\\1" - This expression will match any string where the first letter is repeated every other time.

"(.)(.)(.).*\\3\\2\\1" - This expression matches any string where the first three characters and the last three are the same but in reverse order.


Finally, we will construct regular expressions to match words that meet the descriptions below:

Start and end with the same character. - "^(.).*\\1$"

Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.) - "(..)\\1"

Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.) - "(.)\\1{2,}"

## Conclusions / Findings and Recommendations
Normalizing data frames is definitely a must do when preparing to analyze data. It might feel odd in the beginning taking one table and creating various with relationships, but, although it is not aesthetically pleasing, it makes a huge difference in the feasibility of analyzing certain data sets. Along with normalization, the ability to fetch data through regular expressions is also a must have for data analysis. I look forward to using these methods in future data analysis.

### Work Cited
Casselman, B. (2017, September 28). The economic guide to picking a college major. FiveThirtyEight. https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/

Decomplexify. (2021, November 21). Learn Database Normalization - 1NF, 2NF, 3NF, 4NF, 5NF [Video]. YouTube. https://www.youtube.com/watch?v=GFQaEYEc8_8